---
title: "Getting Started with ivgls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ivgls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(42)
```

## Overview

The **ivgls** package implements three instrumental variable (IV) estimators
for causal variable selection when exposures have network structure:

| Estimator  | Graph penalty | Invalid-IV correction |
|------------|:------------:|:---------------------:|
| `iv_lasso` | No           | No                    |
| `ivgl`     | Yes          | No                    |
| `ivgl_s`   | Yes          | Yes                   |

## Installation

```{r, eval=FALSE}
# glmgraph must be installed first (not on CRAN)
devtools::install_github("cran/glmgraph")

# then install ivgls
install.packages("ivgls")
```

## Quick Start

### 1. Build a graph and Laplacian

```{r}
library(ivgls)

A <- make_graph(p = 20, type = "chain")
L <- get_laplacian(A)
```

### 2. Simulate data

```{r}
bobj <- generate_beta(A, s2 = 4, signal = 2, pattern = "smooth")
dat  <- generate_data(n = 120, p = 20, q = 60,
                      s_alpha = 5, alpha_strength = 3,
                      beta_true = bobj$beta_true)
```

### 3. Fit the estimators

```{r, eval=FALSE}
beta_ivl  <- iv_lasso(dat$Y, dat$X, dat$Z)
beta_ivgl <- ivgl(dat$Y, dat$X, dat$Z, L)
fit_s     <- ivgl_s(dat$Y, dat$X, dat$Z, L, max_iter = 10)
```

### 4. Evaluate recovery

```{r, eval=FALSE}
get_mcc(bobj$active_set, which(abs(beta_ivl)  > 1e-4), p = 20)
get_mcc(bobj$active_set, which(abs(beta_ivgl) > 1e-4), p = 20)
get_mcc(bobj$active_set, which(abs(fit_s$beta)> 1e-4), p = 20)
```

## Running a Simulation Study

```{r, eval=FALSE}
res <- run_simulation(
  B              = 50,
  n              = 100,
  p              = 20,
  q              = 60,
  graph_type     = "chain",
  signal_pattern = "smooth",
  s2             = 4,
  signal         = 2,
  s_alpha        = 5,
  alpha_strength = 3
)

aggregate(cbind(MCC, MSE) ~ Method, data = res, FUN = mean)
```

## Invalid Instrument Detection

When instruments violate the exclusion restriction, `ivgl_s()` identifies
them via the `alpha` component — non-zero entries are invalid instruments.

```{r, eval=FALSE}
fit <- ivgl_s(dat$Y, dat$X, dat$Z, L)
which(abs(fit$alpha) > 1e-4)  # detected invalid instruments
```
